Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study in India

### 2025-08-29 manuscript.Rmd
Work on revision

Trauma, defined as a clinical entity composed of physical injury and the associated response of the body, causes 4.3 million deaths every year [@injuries2020]. Several trauma life support training programmes have been developed to improve the early management of patients in hospitals by providing a structured framework for assessment and treatment [@Mohammad2013; @Jayaraman2014; @Kadhum2020].

The proprietary Advanced Trauma Life Support (ATLS) and low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018]. Observational studies indicate that these programmes may improve patient outcomes [@vestrup_impact_1988; @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016; @kamau_impact_2024; @kumar_evaluating_2025], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].

Several studies, including at least two randomised studies [@Ali1996; @Ali1999], have shown that ATLS is associated with improved knowledge and skills among providers [@Mohammad2013]. Observational evidence also suggests that PTC also leads to improved provider skills [@Kadhum2020]. The missing link is whether this improved knowledge and skill set translates into measurably improved patient outcomes.

Systematic reviews have called for controlled trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020], as conducting such effectiveness trials in settings where they are part of the standard of care is not feasible. Many of these settings are in low- and middle income countries, where trial logistics can be more challenging.

We therefore conducted a pilot study with the aim to assess the feasibility of a cluster randomised controlled trial comparing the effects of ATLS and PTC with standard care on outcomes in adult trauma patients.

@@

There are several significant limitations of this pilot study and, therefore, additional lessons to be learned and factored into the design of a full-scale trial. First, the patient volumes at some of the participating hospitals were lower than expected. A careful assessment of patient volumes as part of the screening process should be included for a full-scale trial. Second, data on complications and causes of death were almost universally missing. Collecting data on these variables will require alternative methods, as these were not explicitly described in the medical records and autopsy reports were not readily available. Third, we did not collect detailed data on the standard care at each hospital. These data should be collected as part of the screening process for a full-scale trial. Fourth, we used sealed envelopes for randomisation, which increases the risk of bias and errors. A full-scale trial should use a computer-generated randomisation system. Fifth, we did not blind the data analysts, but recommend doing so in a full-scale trial. Sixth, we assessed a large number of potential outcomes, and a full-scale trial should focus on the most relevant outcomes. Seventh, the follow-up period was only one month, and changes in some of the outcomes may take longer than that to manifest. The effect on the outcomes is also likely to depend on the adherence to the training, which is not assessed in this pilot study. Finally, owing to a data uploading error, limited data were available from two clusters. At the time of data collection, network and other technical issues were present in some of the clusters. This is a major concern that must be mitigated in a full-scale trial by using a more robust data collection system with local offline backups and careful centralised monitoring.

@@

Previous studies on the effects of ATLS or PTC training on patient outcomes have been observational or quasi-experimental without a control group, with heterogeneous results [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016]. Most suggest that these programmes are associated with improved outcomes, although not all report significant effects [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @kamau_impact_2024; @kumar_evaluating_2025]. In contrast, some studies have shown potential associations with increased mortality [@drimousis_advanced_2011; @CioèPeña2016]. We observed fewer deaths in the intervention arms than in the standard care arm. This difference may have resulted from the randomisation process with a small number of heterogeneous clusters, highlighting the importance of taking varying cluster sizes into account when designing a full-scale trial.

@@

Our study represents the first published attempt to pilot a controlled trial evaluating the effect of trauma life support training on patient outcomes. We conclude that a full-scale cluster randomised trial is feasible after incorporating the lessons of this pilot study.

@@

The code for the analysis is released publicly. The final anonymised dataset is available from the corresponding author upon request.

>>>>

Trauma, defined as a clinical entity composed of physical injury and the associated response of the body, causes 4.3 million deaths every year [@gbd_2021_collaborators_injuries_2024]. Several trauma life support training programmes have been developed to improve the early management of patients in hospitals by providing a structured framework for assessment and treatment [@Mohammad2013; @Jayaraman2014; @Kadhum2020].

Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC) are the most established trauma life support training programmes [@acsAtls2018; @ptcfPtc2018]. Both programmes improve provider knowledge and skills [@Mohammad2013; @Kadhum2020], and while most observational studies associate them with reduced mortality [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @kamau_impact_2024; @kumar_evaluating_2025], some report associations with increased mortality [@drimousis_advanced_2011; @CioèPeña2016].

There are no controlled trials of the effect of ATLS or PTC on patient outcomes, and no trials evaluating any trauma life support training programme in a general trauma population [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024], although a recent cluster randomised trial found that the rural trauma team development course reduces mortality in patients with motorcycle injuries [@lule_impact_2025]. Systematic reviews call for controlled trials of ATLS and PTC in settings where these programmes are not standard care [@Mohammad2013; @Jayaraman2014; @Kadhum2020].

Large scale cluster randomised trials can be complex and challenging, so before embarking on a full scale trial, we conducted a pilot study with the aim to assess the feasibility of a cluster randomised controlled trial comparing the effects of ATLS and PTC with standard care on outcomes in adult trauma patients.

@@

There are several significant limitations of this pilot study and, therefore, additional lessons to be learned and factored into the design of a full-scale trial. First, the patient volumes at some of the participating hospitals were lower than expected. A careful assessment of patient volumes as part of the screening process should be included for a full-scale trial. Second, data on complications and causes of death were almost universally missing. Collecting data on these variables will require alternative methods, as these were not explicitly described in the medical records and autopsy reports were not readily available. Third, we did not collect detailed data on the standard care at each hospital. These data should be collected as part of the screening process for a full-scale trial. Fourth, we used sealed envelopes for randomisation, which increases the risk of bias and errors. A full-scale trial should use a computer-generated randomisation system. Fifth, we did not blind the data analysts, but recommend doing so in a full-scale trial. Sixth, we assessed a large number of potential outcomes, and a full-scale trial should focus on the most relevant outcomes. Seventh, the follow-up period was only one month, and changes in some of the outcomes may take longer than that to manifest. The effect on the outcomes is also likely to depend on the adherence to the training, which is not assessed in this pilot study. Finally, owing to a data uploading error, limited data were available from two clusters. At the time of data collection, network and other technical issues were present in some of the clusters, which could explain this error. Regardless, this is a major concern that must be mitigated in a full-scale trial by using a more robust data collection system with local offline backups and careful centralised monitoring.

@@

Previous studies on the effects of ATLS or PTC training on patient outcomes have been observational or quasi-experimental without a control group, with heterogeneous results [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016; @kamau_impact_2024; @kumar_evaluating_2025]. Most suggest that these programmes are associated with improved outcomes, although not all report significant effects [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @kamau_impact_2024; @kumar_evaluating_2025]. In contrast, some studies have shown potential associations with increased mortality [@drimousis_advanced_2011; @CioèPeña2016]. We observed fewer deaths in the intervention arms than in the standard care arm. This difference may have resulted from the randomisation process with a small number of heterogeneous clusters, highlighting the importance of taking varying cluster sizes into account when designing a full-scale trial.

@@

Our study represents the first published attempt to pilot a controlled trial evaluating the effect of ATLS and PTC on patient outcomes. We conclude that a full-scale cluster randomised trial is feasible after incorporating the lessons of this pilot study.

@@

The code for the analysis is released publicly on GitHub (https://github.com/martingerdin/tern-pilot). The final anonymised dataset is available from the corresponding author upon request.
